\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-16 NAR_DRAFT.tex
change texttt to textrm for consistency

\subsection{The likelihood function}
We designed \batcave to be used as a post-calling evaluation tool with MuTect.
MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in equation \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

@@

The log posterior odds is the log likelihood ratio (\texttt{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as

@@

\subsection{Tumor simulations.}
We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (version 2), downloaded from the COSMIC website.
Each simulated profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
Variants were selected from a combination of the Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:

@@

\subsection{Real Data}
We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
We downloaded the normal and primary tumor bam files generated for the whole genome sequences in Griffith et al., 2015~\citep{Griffith2015} from dbGaP accession number phs000159.v8.p4.
Griffith et al., 2015~\citep{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
The whole exome sequencing bam files generated by Shi et al., 2015~\citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
Shi et al., 2018~\citep{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.

>>>>

\subsection{The likelihood function}
We designed \batcave to be used as a post-calling evaluation tool with MuTect.
MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in equation \ref{eqn:hypothesis} as either \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

@@

The log posterior odds is the log likelihood ratio (\textrm{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as

@@

\subsection{Tumor simulations.}
We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
Tumors were simulated with three different mutation profiles derived from combinations of COSMIC mutation signatures (version 2), downloaded from the COSMIC website.
Each simulated profile includes a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
Variants were selected from a combination of the Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were spiked to create tumors with Bamsurgeon with default parameters~\citep{Ewing2015a},
and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:

@@

\subsection{Real Data}
We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer~\cite{Griffith2015,Shi2018}.
We downloaded the normal and primary tumor bam files generated for the whole genome sequences in Griffith et al., 2015~\citep{Griffith2015} from dbGaP accession number phs000159.v8.p4.
Griffith et al., 2015~\citep{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
The whole exome sequencing bam files generated by Shi et al., 2015~\citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
Shi et al., 2018~\citep{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.
